---
title: "Basic Plotting in R"
author: "Ben Dickins"
date: "`r lubridate::today()`"
output:
  github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparation

## Loading Libraries
Before we can do today's exercises we need to do something extra. We need to get access to some extra functions that aren't available in "base" R. To do this we will load a package into R. In R packages are called libraries and a library is a collection of functions that are usually designed to do similar types of things (e.g., analyzing data in a particular way). In this case we are going to load a really "meta"-package (a package so big that it actually contains other packages) and this is called the `tidyverse`.

```{r tidy-verse}
library(tidyverse)
```

Now don't worry about the "conflicts" message - this isn't an error. (TL;DR: all this means is that some functions in the tidyverse are going to take precedence over functions with the same name in base R.). You can find out more about the tidyverse at [their webpage](https://www.tidyverse.org).


## Setup
Now you've got two tasks:

1. Get your data in the right place. This means downloading the file named "phosphate_assay.tsv" and moving it to a suitable folder.
2. Getting R to look in that place. For this one you will need to "set the working directory".

I will remind you how to do #s 1 and 2 in my video.


## Read in and View Data
Now that we are in the right place, let's read in the data, but this time we are going to use a function from the tidyverse that works with tab-separated values (files ending `.tsv`):
```{r phosphate}
phosphate <- read_tsv("data/phosphate-assay.tsv")
phosphate
```
I will tell you a little bit about that object (which is called a "tibble") in the video and show you a few tricks to view the data in different ways.


# Time to Draw Graphs and Analyze!

## Simple Plot
Did you notice how each replicate is a separate column? Let's start by drawing a graph with the concentration column (on the x axis) and the first replicate (on the y axis):

```{r simple-plot}
plot(x = phosphate$conc_uM, y = phosphate$rep1)
```
It looks like when the concentration goes up so does the value of the replicate (measured as optical density at 600nm).

## Decorate the Plot
Maybe we should put those units into the plot properly:
```{r decorated-plot}
plot(x = phosphate$conc_uM, y = phosphate$rep1, xlab = "concentration (μM)", ylab = "OD (600nm)")
```

## Simple Regression
Let's look at that relationship using a simple linear (regression) model. The function for linear models is `lm()`. In this case we will use this function to construct a model and assign the result to an object called `model1`. We write the name of the model to view its key properties:

```{r model-1}
model1 <- lm(rep1 ~ conc_uM, data=phosphate)
model1
```

So you can see that the slope or gradient is `0.0002288` while the y intercept is `0.1451905`. You can find out much more about the model using `summary(model1)`. This shows the R^2 value and carries out a significance test. (We won't go into this now).

## Adding Regression Line to Plot

```{r regression-plot}
plot(x = phosphate$conc_uM, y = phosphate$rep1, xlab = "", ylab = "")
abline(model1, col="blue")
```
Cool right? We added the regression line to the plot. Did you notice how we added empty information to the plot labels? Try typing this new line of code:

`title(xlab = "concentration (μM)", ylab = "OD (600nm)")`

What happens? Like the `abline()` function above, the `title()` function can be used to decorate a plot after you have drawn it.


## Five Replicates
Now the above is all very well, but we've only plotted one replicate. Just for a second let's remind ourselves of the column names in our data object:

```{r names}
names(phosphate)
```

Let's take the names from the 3rd to the last column and assign them to a new object:
```{r repcols}
repcols <- names(phosphate)[3:7]
repcols
```

Now you can see we need to plot all five columns in repcols.

## Plotting Replicates
Now this is a bit more complicated. Here you'll see two more functions that can be used to add elements to a plot (`points()` and `legend()`) - see if you can figure out what they do:

```{r allrep-plot}
plot(phosphate$conc_uM, phosphate$rep1, pch = 1, ylim = c(0, 0.35), xlab = "", ylab = "")
points(phosphate$conc_uM, phosphate$rep2, pch = 2, xlab = "", ylab = "")
points(phosphate$conc_uM, phosphate$rep3, pch = 3, xlab = "", ylab = "")
points(phosphate$conc_uM, phosphate$rep4, pch = 4, xlab = "", ylab = "")
points(phosphate$conc_uM, phosphate$rep5, pch = 5, xlab = "", ylab = "")
title(xlab = "concentration (μM)", ylab = "OD at 600nm")
legend("topright", legend = repcols, pch = 1:5)
```
To understand the functions you can use the `?` notation for example read the help file for `?points` and see if you can find out what the `pch` argument is doing...

# Plotting in Tidyverse
Tidyverse is a bit of a weird name isn't it? Well it's actually a really good name because it's all about how data is organised. Looking at the data in the phosphate object we can see that each row contains five y values (one for each replicate), but in "tidy" data we expect each observation to be on a separate row and each column to be a single variable. For more detail on this see [this reference](https://r4ds.had.co.nz/tidy-data.html). 

## Tidy Data
How can we get the `phosphate` data into a tidy format? We should have one column for optical density (the y value) and, because every measurement should have its own row, we also need a column to tell us what replicate each measurement is. There is a special helper function in `tidyverse` to help us do this and that is `pivot_longer()`:

```{r phosphate_long}
phosphate_long <- pivot_longer(phosphate, rep1:rep5, names_to = "replicate", values_to = "OD600nm")
phosphate_long
```
Look at the "arguments" (separated by commas) inside `pivot_longer()`. First we pass it the object containing the data, then we tell it which columns contain the data we want to transform/pivot, then we use `names_to` to create a new column that contains these old column names. Finally we use `values_to` to create the _single_ column containing the data. Learn more at the above reference or using `?pivot_longer`. Take a good look at the **tidy** data in our new `phosphate_long` object.


## Plot with "Grammar of Graphics"
Tidyverse contains an awesome function called `ggplot()`. It can be a little hard to learn at first (you may find this [reference chapter](https://r-graphics.org/chapter-ggplot2) helpful), but here is how we can draw a plot with all the data in it:
```{r allreps-ggplot}
ggplot(phosphate_long, aes(x=conc_uM, y=OD600nm, shape=replicate)) + geom_point()
```
First we tell it the name of the data. Then the next argument uses the `aes()` function to tell `ggplot()` the "aesthetic mappings", i.e., which variables are associated with which features of the plot. That's kind of simple for the x and y axes, but try replacing `shape` in the above code with `colour`. What do you see? What if you remove that argument altogether.

Then there is this slightly weird (and unique to `ggplot()`) feature where we add the geometric function that is used to make the plot (in this case `geom_point`). Keeping the `colour` change you made above, try replacing `geom_point()` with `geom_line()`. What do you see? Does that make sense? (It may be pretty but the answer is no! You can't _interpolate_ in that way!)

## Add Regression Line
In `ggplot()` this is really each to do using the `stat_smooth()` function. We just need to add one more thing to the plot (and you might notice that we put the functions on different lines here which is fine as long as there is a plus on every line if we expect more functions):

```{r best-ggplot, fig.cap = "Our most informative figure"}
ggplot(phosphate_long, aes(x=conc_uM, y=OD600nm)) + 
  geom_point(aes(colour=replicate)) + 
  stat_smooth(method=lm)

```
The only thing you might notice here if you look really closely is that we moved the colour attribute into the `geom_points()` function. Try moving it back to the main `ggplot()` call. What happens?

## Plot with Average Values
This is actually very complex, so please don't worry if you don't understand this last example which introduces an idea called the pipe `%>%`, but it shows you how you can summarize data and send it into the next function (and ultimately into `ggplot()`):
```{r mean-ggplot}
phosphate_long %>% 
  group_by(sample, conc_uM) %>% 
  summarise(meanOD600nm = mean(OD600nm)) %>% 
  ggplot(aes(x=conc_uM, y=meanOD600nm)) + geom_point() + stat_smooth(method=lm)
```
Generally you won't need a complex example like this (instead see Figure \\@ref{fig:best-ggplot} which has the advantage that we can see all the data)...

# References
These references were linked above:

* Chang, W., 2018. R graphics cookbook: practical recipes for visualizing data. O'Reilly Media. 2nd edition

* Wickham, H. and Grolemund, G., 2016. R for data science: import, tidy, transform, visualize, and model data. " O'Reilly Media, Inc.".

But when attributing, we mustn't forget about software! R actually contains a function that provides us with bibliographic details! For all of R:
```{r citations-1}
citation()
```

And for a particular package:
```{r citations-2}
citation("tidyverse")
```

# Session Information
For reproducibility we sometimes print out information about our setup and computer:
```{r session-info}
sessionInfo()
```