---
title: "Simple Codes"
subtitle: "Arithmetic and Storing Values"
author: "Ben Dickins"
format: 
  revealjs:
    theme: [moon, custom.scss]
execute:
  echo: true
editor: visual
---

## Expression

There are three types of expression in R\
1. Numbers: 1, 2, 0.2, -5, etc\
2. Strings: alphabets or anything that is input by "" into R\
3. Logical: `TRUE`/`FALSE`

## Expression

-   Numbers:

```{r}
#| eval: false
2+3
```

-   Strings:

```{r}
#| eval: false
"Hello"
```

-   `TRUE`/`FALSE`

```{r}
#| eval: false
3<4
2+4 == 4
```

## Arithmetic Operators

`x + y` sum\
`x - y` subtract\
`x * y` multiply\
`x / y` divide\
`x ^ y` power

## Arithmetic Operators

Do it yourself:

```{r}
2+4
2^3
850/10
220-20
```

## Storing Values

-   In all programming languages we can store values in variables and access them later.
-   This can be done in various ways using a selection of assignment operators.
-   The most commonly used assignment operator is `<-`, see the example below:

```{r}
x <- 3
y <- "Hey!"
```

Note: In most contexts the `=` operator can be used as an alternative.
